---
title: "Applying mixOmics Diablo to the Chronic Lymphocytic Leukemia cohort"
author: 
- name: "Paul-Theodor Pyl"
  affiliation: "SciLifeLab, NBIS, Sweden"
  email: "paul-theodor.pyl@scilifelab.se"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
---

**This document is an attempt to replicate the analysis form the one linked on the MOFA webpage using mixOmics Diablo approach instead. Find the original here: https://raw.githack.com/bioFAM/MOFA2_tutorials/master/R_tutorials/CLL.html**

# Introduction
This vignette shows how to use MOFA+ on the bulk multi-omics data set that was used in the [first publication of MOFA](https://msb.embopress.org/cgi/doi/10.15252/msb.20178124) and the original vignette of the [MOFA package](https://bioconductor.org/packages/release/bioc/vignettes/MOFA/inst/doc/MOFA_example_CLL.html).  

Briefly, the data consists of four omics including DNA methylation, RNA-seq, somatic mutations and drug response data from blood for N=200 patients with Chronic Lymphocytic Leukemia (CLL). The data set is explained in detail [in this article](https://www.jci.org/articles/view/93801) and is publicly available [here](http://bioconductor.org/packages/release/data/experiment/html/BloodCancerMultiOmics2017.html)
  
# Load libraries and data

```{r, message=FALSE}
library(mixOmics)
library(MOFAdata)
library(data.table)
library(ggplot2)
library(tidyverse)
library(utils)
library(MASS)
```

Data is stored as a list of matrices. Features are stored in the rows and samples in the columns
```{r}
utils::data("CLL_data")       
lapply(CLL_data,dim)
```

Sample metadata are stored as a data.frame. Important columns are:  

- **Gender**: m (male), f (female)
- **Age**: age in years
- **TTT**: time (in years) which passed from taking the sample to the next treatment
- **TTD**: time (in years) which passed from taking the sample to patients' death
- **treatedAfter**: (TRUE/FALSE)
- **Died**: whether the patient died (TRUE/FALSE)

The full meta data can be obtained from the Bioconductor package `BloodCancerMultiOmics2017` as `data("patmeta")`.
```{r}
CLL_metadata <- fread("ftp://ftp.ebi.ac.uk/pub/databases/mofa/cll_vignette/sample_metadata.txt")
```

# Block PLSA-Da


```{r}
theY <- CLL_metadata |> filter(!is.na(IGHV))
theX <- lapply(CLL_data, function(x) {
  t(x[,theY$sample])
}) # DIABLO wants samples as rows
result.diablo.cll <- block.plsda(X = theX, Y = theY$IGHV, ncomp = 5) # run the method
plotIndiv(result.diablo.cll) # plot the samples
plotVar(result.diablo.cll) # plot the variables
```

```{r}
for(cidx in seq(1,5)){
  plotLoadings(result.diablo.cll, comp = cidx)
}
```

## Add GeneNames to the mRNA loadings to make it easier to interprete
```{r}
mRNA_loadings <- result.diablo.cll$loadings$mRNA |> as_tibble(rownames = "EnsemblGeneID") |> mutate(abs1 = abs(comp1), abs2 = abs(comp2), absRank1 = rank(-abs1), absRank2 = rank(-abs2))
```

```{r}
#BiocManager::install(c("ensembldb", "AnnotationHub"))
library(AnnotationHub)
ah <- AnnotationHub()
ahDb <- query(ah, pattern = c("Homo Sapiens", "EnsDb"))
ensDb <- ah[["AH116860"]]
GeneNames <- mapIds(ensDb, mRNA_loadings$EnsemblGeneID, keytype = "GENEID", column = "GENENAME")
mRNA_loadings |> mutate(GeneName = GeneNames[EnsemblGeneID]) |> dplyr::filter(absRank1 <= 10 | absRank2 <= 10) |> arrange(absRank1, absRank2)
```

## Investigate mutation loading
```{r}
mutations_loadings <- result.diablo.cll$loadings$Mutations |> as_tibble(rownames = "GeneName")
lowestRank <- mutations_loadings |> dplyr::select(starts_with("comp")) |> mutate(across(everything(), ~ rank(-abs(.x)))) |> rowwise() |> mutate(lowestRank = min(c_across(everything()))) |> pull(lowestRank)
mutations_loadings |> mutate(lowestRank = lowestRank) |> dplyr::filter(lowestRank <= 2) |> arrange(lowestRank)
```
