---
title: "breast.TCGA MOFA vs. Diablo"
author: "Paul-Theodor Pyl"
date: "2024-10-28"
output:
  BiocStyle::html_document:
    toc_float: true
    tox_depth: 3
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Assignment

You are to run the code in this document and add some text below the code blocks and results that comments on what is being done and what your interpretation of the visualisations is. You can work in groups.

Please put the name(s) of all students in your group here:

-   John Doe
-   ...

# BRCA TCGA Data

The Cancer Genome Atlas (TCGA) is a comprehensive and coordinated effort to accelerate our understanding of the molecular basis of cancer through the application of genome analysis technologies, including large-scale genome sequencing. TCGA is a joint effort of the National Cancer Institute (NCI) and the National Human Genome Research Institute (NHGRI), which are both part of the National Institutes of Health, U.S. Department of Health and Human Services.

The TCGA breast cancer cohort: [GDC Data Portal](https://portal.gdc.cancer.gov/projects/TCGA-BRCA)

We will use some of this data to illustrate the use case of DIABLO from the `mixOmics` package, the data is included with the R package sowe can just load it in the session without having to download anything extra.

# Diablo

## SetUp

```{r}
library(mixOmics) # import the mixOmics library
data(breast.TCGA) # extract the TCGA data

# use the mirna, mrna and protein expression levels as predictive datasets
# note that each dataset is measured across the same individuals (samples)
X1 <- breast.TCGA$data.train$mirna
X2 <- breast.TCGA$data.train$mrna  
X3 <- breast.TCGA$data.train$protein
X <- list(mirna = X1, mrna = X2, protein = X3)

Y <- breast.TCGA$data.train$subtype # use the subtype as the outcome variable
```

We get three Omics (miRNA, mRNA and Protein) as well as the cancer subtype, which will be out outcome variable.

## Multiblock PLS-DA

First we do the multiblock PLS-DA (`block.plsda`), trying to discriminate the different subtypes from each other based on the three Omics datasets provided.

```{r, fig.width=8}
result.diablo.tcga <- block.plsda(X, Y) # run the method
plotLoadings(result.diablo.tcga, comp = 1)
plotLoadings(result.diablo.tcga, comp = 2)
plotIndiv(result.diablo.tcga) # plot the samples
plotVar(result.diablo.tcga) # plot the variables
```

## Multiblock sPLS-DA (sparse)

`mixOmics` also provides the multiblock sparse PLS-DA function (`block.splsda`), here we need to specify for each component the number of features we are looking for (we will get 2 components here). Note that the number of features to select for each omic and component is given from the DIABLO tutorial and was presumably empirically determined by playing with different values until the desired result occurred. This is in line with the supervised nature of this approach, i.e. the user specifies the groups of samples and the number of features directly.

```{r, fig.width=8}
# set the number of features to use for the X datasets
list.keepX = list(mirna = c(16, 17), mrna = c(18,5), protein = c(5, 5)) 

# run the method
result.sparse.diablo.tcga <-  block.splsda(X, Y, keepX = list.keepX) 

# plot the contributions of each feature to each dimension
plotLoadings(result.sparse.diablo.tcga, comp = 1)
plotLoadings(result.sparse.diablo.tcga, comp = 2)
plotIndiv(result.sparse.diablo.tcga) # plot the samples
plotVar(result.sparse.diablo.tcga) # plot the variables
```

# MOFA on the same Data

```{r}
library(MOFA2)
library(ggplot2)
library(tidyverse)
```

## Building the model and setting up the options

```{r}
MOFAobject <- create_mofa(lapply(X, t))
MOFAobject
plot_data_overview(MOFAobject)
```

```{r}
data_opts <- get_default_data_options(MOFAobject)
data_opts$scale_views <- TRUE
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 6
train_opts <- get_default_training_options(MOFAobject)
train_opts$convergence_mode <- "slow"
train_opts$seed <- 42
```

```{r}
MOFAobject <- prepare_mofa(MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```

```{r}
MOFAobject <- run_mofa(MOFAobject, outfile="MOFA2_breast.TCGA.hdf5", use_basilisk = TRUE)
```

```{r}
all(rownames(breast.TCGA$data.train$protein) == samples_metadata(MOFAobject)$sample)
samples_metadata(MOFAobject) <- samples_metadata(MOFAobject) |> mutate(Subtype = breast.TCGA$data.train$subtype)
tcgaBRCAclinical <- read_tsv("TCGA-BRCA.Clinical.tsv", col_types = list("age_at_index" = "integer"))
tcgaBRCAclinical <- tcgaBRCAclinical |> dplyr::select(case_submitter_id, age_at_index, ethnicity, vital_status, ajcc_pathologic_stage, primary_diagnosis)
tcgaBRCAclinical <- tcgaBRCAclinical |> dplyr::mutate(SampleID = gsub("TCGA-([A-Z0-9]+)-(.*)", "\\2", case_submitter_id)) |> distinct_all()
write_csv(tcgaBRCAclinical, file = "TCGA-BRCA.Clinical.csv")
all(samples_metadata(MOFAobject)$sample %in% tcgaBRCAclinical$SampleID)
samples_metadata(MOFAobject) <- samples_metadata(MOFAobject) |> inner_join(tcgaBRCAclinical, by = join_by(sample == SampleID))
saveRDS(MOFAobject,"MOFA2_breast.TCGA.rds")
```

## Factor level analysis

```{r}
plot_factor_cor(MOFAobject)
```

```{r}
plot_variance_explained(MOFAobject, max_r2=15)
```

```{r}
plot_variance_explained(MOFAobject, plot_total = TRUE)[[2]]
```

### Correlation with covariates

```{r}
correlate_factors_with_covariates(MOFAobject, 
  covariates = c("Subtype", "age_at_index", "ethnicity", "primary_diagnosis", "vital_status", "ajcc_pathologic_stage"), 
  plot="log_pval"
)
```

```{r}
correlate_factors_with_covariates(MOFAobject, 
  covariates = c("age_at_index", "ethnicity", "primary_diagnosis", "vital_status", "ajcc_pathologic_stage"), 
  plot="log_pval"
)
```

### Factor 1 by Subtype

```{r}
plot_factor(MOFAobject, 
  factors = 1, 
  color_by = "Subtype"
)

plot_factor(MOFAobject, 
  factors = c(1,2,5),
  color_by = "Subtype",
  dot_size = 3,        # change dot size
  dodge = TRUE,           # dodge points with different colors
  add_violin = TRUE,      # add violin plots,
  violin_alpha = 0.25  # transparency of violin plots
)
```

```{r}
plot_weights(MOFAobject,
 view = "mrna",
 factor = 1,
 nfeatures = 10,     # Top number of features to highlight
 scale = TRUE           # Scale weights from -1 to 1
)

plot_top_weights(MOFAobject,
 view = "mrna",
 factor = 1,
 nfeatures = 10
)
```

```{r}
plot_weights(MOFAobject,
 view = "protein",
 factor = 1,
 nfeatures = 10,     # Top number of features to highlight
 scale = TRUE           # Scale weights from -1 to 1
)

plot_top_weights(MOFAobject,
 view = "protein",
 factor = 1,
 nfeatures = 10
)
```

### Factors 2 and 5

```{r}
plot_factor(MOFAobject, 
  factors = c(2,5), 
  color_by = "age_at_index"
)
```

```{r}
plot_factor(MOFAobject, 
  factors = c(2,5), 
  color_by = "primary_diagnosis"
)

plot_factor(MOFAobject, 
  factors = c(1,2,5),
  color_by = "primary_diagnosis",
  dot_size = 3,        # change dot size
  dodge = TRUE,           # dodge points with different colors
  add_violin = TRUE,      # add violin plots,
  violin_alpha = 0.25  # transparency of violin plots
)
```
